%!TEX root = ../thesis.tex
% ******************************* Thesis Appendix A ****************************
\chapter{Related publications}

During my graduate studies I have been an author on a number of publications which are related to this work.
These include the following:

\begin{itemize}[noitemsep]

\item Garrison, Erik, Jouni Sirén, Adam M. Novak, Glenn Hickey, Jordan M. Eizenga, Eric T. Dawson, William Jones et al. ``Variation graph toolkit improves read mapping by representing genetic variation in the reference.'' \emph{Nature biotechnology} (2018).

\item Paten, Benedict, Jordan M. Eizenga, Yohei M. Rosen, Adam M. Novak, Erik Garrison, and Glenn Hickey. ``Superbubbles, ultrabubbles, and cacti.'' \emph{Journal of Computational Biology} (2018).

\item Garg, Shilpa, Mikko Rautiainen, Adam M. Novak, Erik Garrison, Richard Durbin, and Tobias Marschall. ``A graph-based approach to diploid genome assembly.'' \emph{Bioinformatics} (2018).

\item Sirén, Jouni, Erik Garrison, Adam M. Novak, Benedict Paten, and Richard Durbin. ``Haplotype-aware graph indexes.'' \emph{arXiv preprint arXiv:1805.03834} (2018).

\item Paten, Benedict, Adam M. Novak, Jordan M. Eizenga, and Erik Garrison. ``Genome graphs and the evolution of genome inference.'' \emph{Genome Research} (2017).

\item Novak, Adam M., Glenn Hickey, Erik Garrison, Sean Blum, Abram Connelly, Alexander Dilthey, Jordan Eizenga et al. ``Genome graphs.'' \emph{bioRxiv} (2017).

\item Computational pan-genomics consortium. ``Computational pan-genomics: status, promises and challenges.'' \emph{Briefings in Bioinformatics} (2016).

\item Novak, Adam M., Erik Garrison, and Benedict Paten. ``A graph extension of the positional Burrows–Wheeler transform and its applications.'' \emph{Algorithms for Molecular Biology} (2017).

\end{itemize}
